HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-01 submission.Rmd
.

Our project takes this model a step further. While we similarly predict man- or zone coverage when the teams are set, we further leverage the additional information of pre-snap player movements. Using a hidden Markov model (HMM), we model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. Incorporating summary statistics of the probabilistic HMM results as covariates into the existing pre-motion model significantly improves both the AUC and detection accuracy and further allows for evaluating the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.

@@

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as covariates and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder?","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further covariates. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as covariates. An outlet of this problem is to employ suitable summary statistics (`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus","orange")`). In particular, we calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$. From this, we count the number of state switches and sum these numbers across defenders as a first, simple, summary statistics and further calculate across defenders $j = 1, \ldots, 5$ the mean entropy in offensive play $n = 1,\ldots, N$, i.e.:

@@

One limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by pre-snap motion.

>>>>

Our project takes this model a step further. While we similarly predict man- or zone coverage when the teams are set before snaps, we further leverage the additional information of pre-snap player movements. Using a hidden Markov model (HMM), we model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. Incorporating summary statistics of the probabilistic HMM results as features into the existing pre-motion model significantly improves both the AUC and detection accuracy and further allows for evaluating the effectiveness of pre-snap motion in uncovering defensive strategies, providing real-time tactical insights for coaches.

@@

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder?","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as features. An outlet of this problem is to employ suitable summary statistics (`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus","orange")`). In particular, we calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$. From this, we count the number of state switches and sum these numbers across defenders as a first, simple, summary statistics and further calculate across defenders $j = 1, \ldots, 5$ the mean entropy in offensive play $n = 1,\ldots, N$, i.e.:

@@

While we were clearly able to show that motion enables the defensive coverage scheme, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by our extention.
